Estimating the incidence of SARS-CoV-2 from prevalence data

### 2023-05-26 cis-perfect-testing.tex
Keep evolving text

This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (ie: no misclassification bias).
I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
Through a simulation study, I show the applicability of these methods.
However, application in reality must account for misclassification bias, the subject of the following chapter.

Two major challenges arise when dealing with this data: double interval
censoring and left truncation. Double interval censoring arises because
we can only bound the start and end of an infection by observing an
individual's test results transitioning from negative to positive and
vice versa. Left truncation occurs because an individual is infected at
time $t$ is only observed if their infection duration is at least
$t^N - t$ days long, where $t^N$ is the time of their next test.
\todo[inline]{Add figures demonstrating the censoring and truncation here}

To address these challenges, we employ two distinct approaches: the
``individual'' model and the ``total'' model.
\todo[inline]{\textbf{Should I cut one of the models?}}

\todo[inline]{Need to think about what of the below belongs here in the introduction, and turn these parts into a proper literature review}
\todo[inline]{\textbf{Is it better to have a large literature review here, or to discuss relevant literature in each section (the papers on survival analysis in the model section, the papers on priors in that section, COVID estimates where I use them, etc.)?}}

Unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the
data with unobserved times of infection or any information on the
truncated infections; this is due to the large number of infections
(>4000) making this approach computationally expensive.

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
In this chapter I will assume that the time between infection and becoming detectable is negligible.
The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
An episode is \emph{detected} if they test positive at least once; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).

I take a survival analysis approach to analysing this data.
Survival analysis is the area of statistics which estimate the distribution of times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In this context, the distribution to estimate is that of duration of positivity.

@@

The first issue is double interval censoring.
Explain issue, brief overview of literature.

The second issue is truncation.

These combination of issues are rarely encountered in biomedical data (ref).

@@

\section{Modelling the duration}\label{perf-test:sec:model}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{perf-test:fig:partitionSpace} \todo[inline]{Digitalise figure of regions.}}
\end{figure}

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with
the frequentist likelihood-based approach, up to the priors on other
parameters \cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

@@

The hazard is convenient to use because it is unconstrained except in the interval [0 ,1]; include REFs.
Survival needs to be monotonic.

Discuss literature on Beta Process priors (maybe this should be in a broader literature review part?).

@@

Instead propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.

@@

First, transform ATACCC analsysis into a posterior over survival time.

@@

Finally, use a Beta Process to "discount" some of the information in ATACCC towards a vague prior (as in the previous section).

@@

We require a ground truth for the distribution giving duration of positivity.
The majority of this distribution is based on \autoref{ATACCC}'s estimates from the ATACCC data.
However, the tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
These curves and the combined curve are compared in figure \@ref(fig:duration-curves).
Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.

Sarah uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function (@roystonSTPM; @roystonFlexible) with covariates introduced via proportional odds.

We know of two biases introduced from this analysis: “There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test. Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short). However, there is another bias due to the survey missing positive episodes entirely if they are short. This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes. This will bias the clearance curves upwards (making the estimates too long).”~\autocite{cisMethodsONS}.

@@

\todo[inline]{Figure with duration distributions}

@@

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside of this period, all tests are negative (ie: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).

@@

In the fixed testing schedules, there is no automatic choice of sample size.
We simulate a large number of episodes, and then choose to keep 4800 episodes\todo{Check how many this is} where at least one positive test was returned.

>>>>

This chapter aims to estimate the duration of SARS-CoV-2 PCR positivity from CIS data if perfect testing can be assumed (i.e.: no misclassification bias).
I take a survival analysis approach, considering that neither the start nor end time of positivity are known exactly (double interval censoring) and that infections may be missed (truncation), explained in the following paragraphs.
Through a simulation study, I show the applicability of these methods.
However, application in reality must account for misclassification bias, the subject of the following chapter.

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \autoref{imperf-test}).
The episode's \emph{duration of positivity} is the number of days that they are detectable for; I use time discretised to days and hence this is at least 1 day.
Note that the relationship between positivity and infectiousness is complex\todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
An episode is \emph{detected} when they test positive at least once, if this occurs; episode $i$ is detected at time $t$ if and only if: $i$ is detectable at $t$, $i$ is tested at $t$, and the test returns positive (a false negative does not occur).

I take a survival analysis approach to analysing this data.
Survival analysis is the area of statistics which estimate the distribution of times between events.
It is broadly applicable to many domains, such as the time-to-failure for a mechanical system or the effect of a treatment on time spent in hospital.
In this context, the distribution to estimate is that of duration of positivity.
The \emph{initiating event} is when the duration starts, here becoming detectable.
The \emph{terminating event} is when the duration starts, here becoming no longer detectable.

@@

The first issue is double interval censoring.
An event (either the initiating or terminating) is \emph{interval censored} if its value is only to lie within a range.
\emph{Double interval censoring} means both the initiating and the terminating events are interval censored.
The initiating event is censored because we only observe the change from being not detectable (i.e.: negative) at one test, to detectable (i.e.: positive) at the following test; at the terminating event we see the opposite occurring (see\todo{interval censoring figure}).

Interval censoring is a well-studied area in the biostatistics and broader statistical literature\todo{interval censoring refs}.
Double interval censoring is less frequent, although has been considered previously\todo{double interval censoring refs}.

The second issue is truncation.
Truncation occurs because we do not observe some episodes: an episode can occur between tests and hence the episode is never detected (see figure\todo{truncation figure}).
We do not know how

This combination of issues are rarely encountered in biomedical data\todo{refs to combining double interval censoring with truncation}.

@@

\section{Modelling the duration}\label{perf-test:sec:model}
Now I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster).
Only limited resources are unavailable within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (eg: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this is due to the large number of infections (>4000 in the time period considered) making this approach computationally expensive.

@@

\begin{figure}
\includegraphics[width=\textwidth]{cis-perfect-testing/images/likelihood_areas.jpg}
\caption{Admissible, $\alpha_i$, inadmissible, $\beta_i$ and
untruncated, $\Omega_i^C$ regions for an episode $i$ in an
individual whose first test was at time 0 and negative, with subsequent
negative tests at times 7, 14, and 56, and positive tests at times 21
and 28. Purple region is $\alpha_i$, orange region (which includes all
negative infection times) is $\Omega_i^C$ and unshaded region is
$\beta_i$.\label{perf-test:fig:partitionSpace} \todo[inline]{Digitalise figure of regions.}}
\end{figure}

@@

The prior $p(n_i) \propto 1/n_i$ means that the model coincides with
the frequentist likelihood-based approach, up to the priors on other
parameters~\cites[section 4.2]{dempsterMaximum}{heiseyModelling}[section 8.7.5]{gelmanBayesian}, and is
the reference prior for this quantity\autocite{heBayesiana}. Under this prior,
$\sum_{n_i=1}^\infty n_i p_{it}^{n_i- 1} p(n_i) \propto \sum_{n_i=1}^\infty p_{it}^{n_i-1} = 1/(1-p_{it})$.
% (\protect\hyperlink{ref-dempsterMaximum}{Dempster, Laird, and
% Rubin} (\protect\hyperlink{ref-dempsterMaximum}{1977}) section 4.2;
% \protect\hyperlink{ref-heiseyModelling}{Heisey and Nordheim}
% (\protect\hyperlink{ref-heiseyModelling}{1995});
% \protect\hyperlink{ref-gelmanBayesian}{Gelman et al.}
% (\protect\hyperlink{ref-gelmanBayesian}{2013}) section 8.7.5)

@@

In this section, I consider the various parameterisation of the survival function that are available.
This has implications for the choice of prior  it is not possible to be vague on all sensible parameterisations.
Informative priors might also be attractive, and allow the incorporation of the estimates from \autoref{ATACCC}.

@@

The hazard is convenient to use because it is unconstrained except in the interval [0, 1]\todo{ref where been used previously}.
Survival needs to be monotonic.

The main building basis of the priors I propose is the Beta Process\todo{Beta Process refs}.
The Beta Process is...

@@

Instead, I propose Beta(0.1, 1.9), linking to previous literature for a point estimate and showing vague figures.

@@

First, transform ATACCC analysis into a posterior over survival time.

@@

Finally, use a Beta Process to discount some information in ATACCC towards a vague prior (as in the previous section).

@@

We require a ground truth for the distribution giving duration of positivity.
The majority of this distribution is based on \autoref{ATACCC}'s estimates from the ATACCC data.
However, the tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
ATACCC is thought to estimate accurately the majority of the distribution but not the tail due to a small sample size and model uncertainty.
These curves and the combined curve are compared in \autoref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $d_i$, is then an independent draw from this distribution.

\begin{figure}
  \todo[inline]{Find figure comparing different duration distributions}
  \caption{Comparison of the duration distributions, see main text for details of how each is derived. \label{perf-test:fig:duration-dist}}
\end{figure}

Sarah uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval-censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.

We know of two biases introduced from this analysis.
\enquote{There is a bias in estimating the clearance distribution because the analysis used to estimate how long a person stays positive only starts from their first positive test.
Since (most) people will have become positive on an earlier day, this will bias the clearance curves downwards (making the estimates too short).
However, there is another bias due to the survey missing positive episodes entirely if they are short.
This means that our dataset has fewer short positive episodes than in the population as a whole, and that the sample used to run the survival analysis is biased towards people with longer positive episodes.
This will bias the clearance curves upwards (making the estimates too long).}~\autocite{cisMethodsONS}.

@@

@@

The first day individual $i$ can be positive is their day of infection, $t_i^\text{onset}$, the last day is $t_i^\text{final} = t_i^\text{onset} + d_i - 1$ where $d_i$ is the duration of positivity for individual $i$.
Outside this period, all tests are negative (ie: no false positives occur). 
Any test between $t_i^\text{onset}$ and $t_i^\text{final}$ (inclusive) is positive with probability $p_{sens}$ (that is, each test is an independent Bernoulli trial).

@@

### 2023-05-26 main.tex
Keep evolving text

\documentclass[a4paper]{report}
\usepackage{graphicx}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{textcomp} % provide symbols
\usepackage{todonotes}
% \usepackage{microtype}

@@

\section{Data}
The CIS continually enrols individuals and tests them on an ongoing basis, following a
specified testing protocol. The protocol specifies that the initial five
tests are spaced every 7 days, then the spacing is every 28 days.
However, real-world considerations often lead to variations in test
times and missed tests. Our analysis focuses solely on the binary result
(positive or negative) of these tests, assuming no misclassification
bias.

The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
A full description of the study can be found in the supplementary materials of [Pouwels et al. 2020](https://www.thelancet.com/journals/lanpub/article/PIIS2468-2667(20)30282-6/fulltext#supplementaryMaterial) or [the study protocol](https://www.ndm.ox.ac.uk/covid-19/covid-19-infection-survey/protocol-and-information-sheets).
In brief, households are invited to the study from databases held by the ONS.
The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
Within a household, all indvidiuals aged 2 and over were invited to participate.
Once invited, an enrolement swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolement) after which monthly swabs are taken.

\todo[inline]{Add some descriptive figures eg those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}

@@

>>>>

\documentclass[a4paper]{report}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[style=british]{csquotes}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{textcomp} % provide symbols
\usepackage{todonotes}
% \usepackage{microtype}

@@

\chapter{Duration from biomarkers} \label{ATACCC}

@@

\appendix

Random bits of text that I've written that don't currently fit anywhere but will likely be useful in future

\section{CIS study description}
The CIS continually enrols individuals and tests them on an ongoing basis, following a
specified testing protocol. The protocol specifies that the initial five
tests are spaced every 7 days, then the spacing is every 28 days.
However, real-world considerations often lead to variations in test
times and missed tests. Our analysis focuses solely on the binary result
(positive or negative) of these tests, assuming no misclassification
bias.

The CIS (Coronavirus Infection Survey) is a longitudinal study on a representative sample of households.
A full description of the study can be found in the supplementary materials of \textcite{pouwelsCommunity} or the study protocol~\autoref{cisProtocol}.
In brief, households are invited to the study from databases held by the ONS.
The number invited was to satisfy various target number of individuals swabbing per fortnight; relevant to the discussion in this document is that the target increased from 2,500 by the end of May 2020 to 150,000 in October 2020.
Within a household, all individuals aged 2 and over were invited to participate.
Once invited, an enrolment swab would be taken followed by 4 further weekly swabs (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which monthly swabs are taken.

\todo[inline]{Add some descriptive figures e.g. those in ~/COVID/ons-incidence/duration_estimation/reproducible-sims/reports/simulation-studies.pdf}
